Score-based tests of measurement invariance: Use in practice

### 2014-07-14 frontiers/IMPSpres-knitr.Rnw
update slide 10 three matrix.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@357 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{frame}
\frametitle{Possible Solution}
\begin{itemize}
\item Likelihood Ratio Test (LRT): Require estimates under null
  hypothesis and alternative.
%  \begin{itemize}
%    \item Ignore the nature of ordinal variable, treat it as
%      categorical varaible. 
%    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
%    \item Rather tedious and time consuming to fit and compare all possible
%      models. e.g, for testing one parameter, you need to fit $2^{m-1}$ 
%      models, $m$ is the level of ordinal variable.
%    \end{itemize}
\item Wald Test: Require estimates under alternative. 
\item Lagrange Multiplier Test (Score Test): Require estimates under
  null hypotheis.
\item \textbf{Score Test generalizations w.r.t. general auxiliary
    variable (categorical, ordinal, or continuous)}. 
\end{itemize}
\end{frame}

@@

\section{Theoretical Background}
\begin{frame}
  \frametitle{Score-based Tests}
 \begin{itemize}
  %  \item The score-based tests \cite{MerZei13, MerFanZei}
  %    rely on first derivatives of
  %    the model's log-likelihood function.
    \item Access deviations in the model score \cite{MerZei13,
      MerFanZei}, i.e., the first derivatives of the model's log
      likelihoood function.  
    \item We consider individual terms ({\em scores}) of the gradient.  
      These scores tell us how well a particular parameter describes a 
      particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}

@@

\begin{frame}
  \frametitle{Visual Explanation of the Score-based Test}
  \begin{columns}
    \begin{column}{0.33\textwidth}     
      \begin{block}{Unordered Score}
        \resizebox{3.3cm}{!}{ 
          $
            \begin{blockarray}{lccccc}
              \text{obs} &\theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & \text{age} \\
               \begin{block}{l(cccc)c}
                 &4.12 & 3.48 & \ldots & -3.11 & 11.3\\
                 &1.32  & -4.32 & \ldots & -2.34 & 10.5\\
                 &3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
                 &-5.78 & 3.50 & \ldots & 7.67 & 18.9\\
                 &\vdots & \vdots & \vdots & \vdots & 17.1\\ 
                 &12.34 & 13.32 & \ldots & 6.89 & 18.6\\              
               \end{block}
            \end{blockarray} 
           $
          }
        \end{block}   
     \end{column}

    \begin{column}{0.33\textwidth}     
      \begin{block}{Ordered Score}
        \resizebox{3.3cm}{!}{ 
          $
            \begin{blockarray}{lccccc}
              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & \text{age} \\
               \begin{block}{l(cccc)c}
                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5\\
                 x_{2} &  3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
                 x_{3} &  4.12 & 3.48 & \ldots & -3.11 & 11.3\\
                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1\\ 
                 x_{n-1} &  12.34 & 13.32 & \ldots & 6.89 & 18.6\\
                 x_{n} & -5.78 & 3.50 & \ldots & 7.67 & 18.9\\
               \end{block}
            \end{blockarray} 
           $
          }
        \end{block}   
     \end{column}

     \begin{column}{0.33\textwidth}
          \begin{block}{Cumulative Score}
          \resizebox{3.3cm}{!}{
           $
            \begin{blockarray}{lccccc}
              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & \text{age} \\
               \begin{block}{l(cccc)c}
                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
                 x_{2} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
                 x_{3} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1 \\
                 x_{n-1} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
                 x_{n}& 0     &   0     & \ldots & 0 & 18.9 \\
               \end{block}
            \end{blockarray} 
           $
         }  
        \end{block}
      \end{column}
   \end{columns}
\end{frame}

@@

\begin{frame}
   \frametitle{Test Statistics}
\begin{itemize}
    \itemsep8pt
  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equvialent to LRT).
\item Ordinal $V$: Weighted double maximum test ($\mathit{WDM}_o$);\\
Maximum Lagrange multiplier test ($\max\mathit{LM}_o$).
\item Continuous $V$: Double maximum test ($\mathit{DM}$)\\
Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
maximum Lagrange multiplier ($\max\mathit{LM}$)\\
\end{itemize}
\end{frame}

>>>>

\begin{frame}
\frametitle{Possible Solution}
\begin{itemize}
\item Likelihood Ratio Test (LRT): Require estimates under null
  hypothesis and alternative.
%  \begin{itemize}
%    \item Ignore the nature of ordinal variable, treat it as
%      categorical varaible. 
%    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
%    \item Rather tedious and time consuming to fit and compare all possible
%      models. e.g, for testing one parameter, you need to fit $2^{m-1}$ 
%      models, $m$ is the level of ordinal variable.
%    \end{itemize}
\item Wald Test: Require estimates under alternative. 
\item Lagrange Multiplier Test (Score Test): Require estimates under
  null hypothesis.
\item \textbf{Score Test generalizations w.r.t. general auxiliary
    variable (categorical, ordinal, or continuous)}. 
\end{itemize}
\end{frame}

@@

\section{Theoretical Background}
\begin{frame}
  \frametitle{Score-based Tests}
 \begin{itemize}
  %  \item The score-based tests \cite{MerZei13, MerFanZei}
  %    rely on first derivatives of
  %    the model's log-likelihood function.
    \item Access deviations in the model score \cite{MerZei13,
      MerFanZei}, i.e., the first derivatives of the model's log
      likelihood function.  
    \item We consider individual terms ({\em scores}) of the gradient.  
      These scores tell us how well a particular parameter describes a 
      particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}

@@

\begin{frame}[t]
  \frametitle{Visual Explanation of the Score-based Test}
  \begin{columns}
    \begin{column}{0.45\textwidth}     
      \begin{block}{\scriptsize{Unordered Score}}
        \resizebox{4.0 cm}{!}{ 
          $
            \begin{blockarray}{lccccc}
              \text{obs} &\theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & V \\
               \begin{block}{l(cccc)c}
                x_{1} &4.12 & 3.48 & \ldots & -3.11 & 11.3\\
                x_{2}  &1.32  & -4.32 & \ldots & -2.34 & 10.5\\
                x_{3} &3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
                x_{..} &\vdots & \vdots & \vdots & \vdots & \vdots\\ 
                x_{n-1} &-5.78 & 3.50 & \ldots & 7.67 & 18.9\\            
                x_{n} &12.34 & 13.32 & \ldots & 6.89 & 18.6\\              
               \end{block}
            \end{blockarray} 
           $
          }
        \end{block}   
     \end{column}

    \begin{column}{0.45\textwidth}     
      \begin{block}{\scriptsize{Ordered Score}}
        \resizebox{4.0 cm}{!}{ 
          $
            \begin{blockarray}{lccccc}
              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & V \\
               \begin{block}{l(cccc)c}
                 x_{(1)} &  1.32  & -4.32 & \ldots & -2.34 & 10.5\\
                 x_{(2)} &  3.37 & 8.32 & \ldots & 63.45 & 11.0\\ 
                 x_{(3)} &  4.12 & 3.48 & \ldots & -3.11 & 11.3\\
                 x_{(..)} & \vdots & \vdots & \vdots & \vdots & \vdots\\ 
                 x_{(n-1)} &  12.34 & 13.32 & \ldots & 6.89 & 18.6\\
                 x_{(n)} & -5.78 & 3.50 & \ldots & 7.67 & 18.9\\
               \end{block}
            \end{blockarray} 
           $
          }
        \end{block}   
     \end{column}
   \end{columns}

   \begin{block}{\scriptsize{Empirical Cumulative Score}}
          $\bm B(\cdot;\hat{\bm \theta}) = \hat{\bm I}^{-1/2}n^{-1/2} \times$
            \resizebox{4.2cm}{!}{
           $
            \begin{blockarray}{lccccc}
               \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & V \\
               \begin{block}{l(cccc)c}
                 x_{(1)} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
                 x_{(2)} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
                 x_{(3)} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
                 x_{(..)} & \vdots & \vdots & \vdots & \vdots & \vdots\\
                 x_{(n-1)} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
                 x_{(n)}& 0     &   0     & \ldots & 0 & 18.9 \\
               \end{block}
            \end{blockarray} 
           $
         }
        \end{block}

@@

\begin{frame}
   \frametitle{Test Statistics}
\begin{itemize}
    \itemsep8pt
  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equivalent to LRT).
\item Ordinal $V$: Weighted double maximum test ($\mathit{WDM}_o$);\\
Maximum Lagrange multiplier test ($\max\mathit{LM}_o$).
\item Continuous $V$: Double maximum test ($\mathit{DM}$)\\
Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
maximum Lagrange multiplier ($\max\mathit{LM}$)\\
\end{itemize}
\end{frame}
